Still Missing a Cool Title

### 2006-05-31 talk/dls2006/paper.tex
it seems we really want pdf
first round add doing real citations

Our current efforts, and the present paper, focus on this tool-suite.
We will not describe the Standard Interpreter component of PyPy in the
sequel, other than mention that it is written in RPython and can thus be
translated.  At close to 90,000 lines of code, it is the largest RPython
program that we have translated so far.  More information can be found
in [S].

@@

But RPython comes with automatic memory management, and this first
transformation step produces flow graphs that also assume automatic
memory management.  Generating C code directly from there produces a
fully leaking program, unless we link it with an external garbage
collector (GC) like the Boehm conservative GC [Boehm], which is a
viable option.

@@

More information about these transformations can be found in [T].

@@

\begin{figure}
\centering
\epsfig{file=image/typesystems.eps, scale=0.667}
\caption{type systems; the arrows are transformations.}
\label{fig_typesystems}
\end{figure}

@@

\begin{figure*}
\centering
\epsfig{file=image/flowlattice.eps, scale=0.65}
\caption{The lattice order of the flow object space}
\label{flowlattice}
\end{figure*}

@@

Note that we produce flow graphs in Static Single Information (SSI)
form, an extension of Static Single Assignment ([SSA]): each variable is
only used in exactly one basic block.  All variables that are not dead
at the end of a basic block are explicitly carried over to the next
block and renamed.

While the Flow Object Space is quite a short piece of code -- its core
functionality holds in 300 lines -- the detail of the interactions
sketched above is not entierely straightforward; we refer the reader to
[D] for more information.

@@

\begin{figure*}
\centering
\epsfig{file=image/lattice1.eps, scale=0.75}
\caption{The overall lattice}
\label{latticeoverall}
\end{figure*}

\begin{figure*}
\centering
\epsfig{file=image/lattice2.eps, scale=0.75}
\caption{The part about instances and nullable instances, assuming a
  simple class hierarchy with only two direct subclasses of
  \texttt{object}.}
\label{latticedetail}
\end{figure*}

@@

Squeak [Squeak] is a Smalltalk implementation in Smalltalk. It uses SLang, a
very restricted subset of Smalltalk with few types and strict
conventions, which can be mostly directly translated to C.
The VM, the object memory and the garbage collector support are
explicitly written together in this style. Again simplicity and
portability were the major goals, as opposed to sophisticated manipulation
and analysis or "weaving" in of features as transformation aspects.

Jikes RVM [DBLP:journals/ibmsj/AlpernABBCCCFGHHLLMNRSSSSSSW00] is a
Java VM and Just-In-Time compiler written in Java.
Bootstrapping happens by self-applying the compiler on a host VM, and
dumping a snapshot from memory of the resulting native code.
This approach directly enables high performance, at the price of
portability -- as usual with pure native code emitting
approaches. Modularity of features, when possible, is achieved with
normal software modularity. The indirection costs are taken care of by
the compiler performing inlining (which is sometimes even explicitly
requested).  In particular this modular approach is used
for implementing a range of choices for GC support [Jikes-GC].  This was
the inspiration for PyPy's own GC framework, although much more tuning
and work went into Jikes RVM.  PyPy's own GC framework also exploits
inlining of helpers and barriers to recover performance.

Jikes RVM's native JIT compilers [Jikes-JIT]
are not meant to be retargetted to run in other environments
that hardware processors, for example in a CLR/.NET
runtime. Also Jikes RVM pays the complexity of writing
a JIT up-front, which also means that features and semantics of the
language are encoded in the JIT compiler code.  Major changes of the
language are likely to correspond to major surgery of the JIT.

@@

Dynamic languages enable the definition of multiple custom type systems,
similar to \textit{pluggable type systems} in [Bracha] but with simple
type inference instead of explicit annotations.  This proved a key
feature in implementing our translation tool-chain, because it makes a
many-levels approach convenient: each abstraction level can provide an
implementation for some of the features that the higher levels
considered primitive.  It offsets the need to define a minimal kernel of
primitives and build everything on top of it; instead, we have been able
to implement, so to speak, the right feature at the right level.

>>>>

Our current efforts, and the present paper, focus on this tool-suite.
We will not describe the Standard Interpreter component of PyPy in the
sequel, other than mention that it is written in RPython and can thus be
translated.  At close to 90,000 lines of code, it is the largest RPython
program that we have translated so far.  More information can be found
in \cite{A}.

@@

But RPython comes with automatic memory management, and this first
transformation step produces flow graphs that also assume automatic
memory management.  Generating C code directly from there produces a
fully leaking program, unless we link it with an external garbage
collector (GC) like the Boehm conservative GC \cite{Boehm}, which is a
viable option.

@@

More information about these transformations can be found in \cite{T}.

@@

\begin{figure}
\centering
\includegraphics[scale=0.667]{image/typesystems.pdf}
\caption{type systems; the arrows are transformations.}
\label{fig_typesystems}
\end{figure}

@@

\begin{figure*}
\centering
\includegraphics[scale=0.75]{image/flowlattice.pdf}
\caption{The lattice order of the flow object space}
\label{flowlattice}
\end{figure*}

@@

Note that we produce flow graphs in Static Single Information (SSI, \cite{SSI})
form, an extension of Static Single Assignment (\cite{SSA}): each variable is
only used in exactly one basic block.  All variables that are not dead
at the end of a basic block are explicitly carried over to the next
block and renamed.

While the Flow Object Space is quite a short piece of code -- its core
functionality holds in 300 lines -- the detail of the interactions
sketched above is not entierely straightforward; we refer the reader to
\cite{D} for more information.

@@

\begin{figure*}
\centering
\includegraphics[scale=0.75]{image/lattice1.pdf}
\caption{The overall lattice}
\label{latticeoverall}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[scale=0.75]{image/lattice2.pdf}
\caption{The part about instances and nullable instances, assuming a
  simple class hierarchy with only two direct subclasses of
  \texttt{object}.}
\label{latticedetail}
\end{figure*}

@@

Squeak \cite{Squeak} is a Smalltalk implementation in Smalltalk. It uses SLang, a
very restricted subset of Smalltalk with few types and strict
conventions, which can be mostly directly translated to C.
The VM, the object memory and the garbage collector support are
explicitly written together in this style. Again simplicity and
portability were the major goals, as opposed to sophisticated manipulation
and analysis or "weaving" in of features as transformation aspects.

Jikes RVM \cite{DBLP:journals/ibmsj/AlpernABBCCCFGHHLLMNRSSSSSSW00} is a
Java VM and Just-In-Time compiler written in Java.
Bootstrapping happens by self-applying the compiler on a host VM, and
dumping a snapshot from memory of the resulting native code.
This approach directly enables high performance, at the price of
portability -- as usual with pure native code emitting
approaches. Modularity of features, when possible, is achieved with
normal software modularity. The indirection costs are taken care of by
the compiler performing inlining (which is sometimes even explicitly
requested).  In particular this modular approach is used
for implementing a range of choices for GC support \cite{JikesGC}.  This was
the inspiration for PyPy's own GC framework, although much more tuning
and work went into Jikes RVM.  PyPy's own GC framework also exploits
inlining of helpers and barriers to recover performance.

Jikes RVM's native JIT compilers \cite{Jikes-JIT}
are not meant to be retargetted to run in other environments
that hardware processors, for example in a CLR/.NET
runtime. Also Jikes RVM pays the complexity of writing
a JIT up-front, which also means that features and semantics of the
language are encoded in the JIT compiler code.  Major changes of the
language are likely to correspond to major surgery of the JIT.

@@

Dynamic languages enable the definition of multiple custom type systems,
similar to \textit{pluggable type systems} in \cite{PlugType} but with simple
type inference instead of explicit annotations.  This proved a key
feature in implementing our translation tool-chain, because it makes a
many-levels approach convenient: each abstraction level can provide an
implementation for some of the features that the higher levels
considered primitive.  It offsets the need to define a minimal kernel of
primitives and build everything on top of it; instead, we have been able
to implement, so to speak, the right feature at the right level.
